---
title: "Comparison Between Methods"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  classoption: landscape
header-includes:
- \usepackage{titling}
- \usepackage{wrapfig}
- \usepackage{lipsum}
- \usepackage{pdflscape}
- \pretitle{\begin{center} \includegraphics[width=2in,height=2in]{1200px-MassDPH_svg.png}\LARGE\\}
- \posttitle{\end{center}}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#Load in Libraries
library(readxl)
library(plyr)
library(dplyr)
library(arsenal)
library(ggplot2)
library(writexl)
library(openxlsx)
library(lubridate) 
library(data.table)
library(kableExtra)
library(tidyr)
library(janitor)
library(scales)
library(kableExtra)
library(tinytex)
library(yaml)
library(ggpubr)
library(cowplot)
library(zoo)
library(formattable)
library(treemap)
library(viridis)
library(paletteer)
library(rlist)
library(magrittr)
library(mosaic)
library(tidyverse)

```

```{r, include = F}

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)

defOut <- knitr::knit_hooks$get("plot")  # save the default plot hook 
knitr::knit_hooks$set(plot = function(x, options) {  # set new plot hook ...
  x <- defOut(x, options)  # first apply the default hook
  if(!is.null(options$wrapfigure)) {  # then, if option wrapfigure is given ...
    # create the new opening string for the wrapfigure environment ...
    wf <- sprintf("\\begin{wrapfigure}{%s}{%g\\textwidth}", options$wrapfigure[[1]], options$wrapfigure[[2]])
    x  <- gsub("\\begin{figure}", wf, x, fixed = T)  # and replace the default one with it.
    x  <- gsub("{figure}", "{wrapfigure}", x, fixed = T)  # also replace the environment ending
  }
  return(x)
})
```
\pagebreak


```{r setup and summary, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

# Pull in Data
Data<-read_excel(here::here("J:/Wastewater Surveillance/Emily/Comparisons/Methods Comparison","KFKitsPoCData_051823.xlsx")) 

Data <- Data %>% drop_na()

# Table
Data2 <- subset(Data, select = c(Method, Accession, `ddPCR date`, `PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`))

knitr::kable(head(Data2, 30), caption = "Data Preview") %>%
  kable_styling("bordered", position = "center") %>%
  kable_styling(latex_options="scale_down")

```

\pagebreak

# Graph of Concentrations

```{r graphs, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

ggplot(Data, aes(x=`Accession`, y=`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`, group = factor(Method))) + 
  labs(title="Found SARS-CoV-2 Concentrations for 3 Different Methods") +  xlab("Sample") + ylab("SARS-CoV-2 Concentration (copies/L)")  + 
  theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
  geom_line(aes(color=Method))+  geom_point(aes(color=Method))+ theme(legend.position='bottom')


```

\pagebreak

# Comparisons - Found Values

We do 2 sets of comparisons, comparing both new methods to the current.

```{r graphs all vs all, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

# Reshape
Data3 <- Data2 %>% select(-`ddPCR date`)
AllvsAll<- Data3 %>% 
  pivot_wider(names_from=c('Method'), 
              values_from='PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)') 

AllvsAllfiltered <- AllvsAll %>% filter(Accession !='23WW361') #this accession is an outlier
AllvsAllfiltered <- AllvsAllfiltered %>% filter(Accession !='23WW291') #low precision
AllvsAllfiltered <- AllvsAllfiltered %>% filter(Accession !='23WW348') #low precision
AllvsAll <- AllvsAllfiltered

ggplot(AllvsAll, aes(x=`Current`, y=`MVP I Kit`)) + 
  labs(title="Current vs MVP I Kit Concentration Measurements - outliers and unsats dropped", subtitle = )+
  ylab("MVP I Kit")  + theme_bw() + 
  geom_smooth(method = "lm", se=FALSE, color="black") + 
  geom_point() + stat_regline_equation(label.x = 100, label.y = 240000)+ stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),label.x = 100, label.y = 225000)


# Plot
ggplot(AllvsAll, aes(x=`Current`, y=`WW Kit`)) + 
  labs(title="Current vs WW Kit Concentration Measurements - outliers and unsats dropped") +  
  xlab("Current Method") + 
  ylab("WW Kit")  + theme_bw() + 
  geom_smooth(method = "lm", se=FALSE, color="black") + 
  geom_point() + stat_regline_equation(label.x = 100, label.y = 300000) +  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), label.x = 100, label.y = 280000)

```

Both kits are found to produce results correlated with those of the current method. With an $R^2$ of 0.74, the both the MVP I and WW Kit concentrations seem to have a great "fit" against the current method. To convert between the methods' found concentrations, an equation is given for the line of best fit on the scatterplot, in which "x" is the x axis of the graph (current method).

\pagebreak

## ANOVA

```{r ANOVA, echo=TRUE, warning=FALSE}

# Model
Data <- Data %>% filter(Accession !='23WW361')
Data <- Data %>% filter(Accession !='23WW291') #low precision
Data <- Data %>% filter(Accession !='23WW348') #low precision
BigModel <- lm(Data$`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)` ~ Data$Method)
summary(BigModel)

view(Data)
```

The model fails to find a significant difference in concentrations between methods.

## Pairwise T Tests

```{r t tests, echo=TRUE, warning=FALSE}

t.test(AllvsAll$`Current`, AllvsAll$`MVP I Kit`, paired = TRUE, alternative = "two.sided")
t.test(AllvsAll$`Current`, AllvsAll$`WW Kit`, paired = TRUE, alternative = "two.sided")

```

The T test finds a significant pair-wise difference between the current and MVP I kit results, but not between the current and the WW kit.

\pagebreak

# Comparisons - LOG TRANSFORM

We repeat, using the log10 values in order to assure normal distribution.

```{r graphs log, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}
hist(Data3$`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`, main="Found concentrations exhibit a right skew", 
     xlab = "Concentration", breaks=20)
hist(log10(Data3$`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`), main="Log transform on concentrations generates more normal distribution", 
     xlab = "Log10(concentration)", breaks=20)

```
The concentrations exhibit a marked right skew. A log10 transform creates a more normal-looking distribution for the data. This will help us to better satisfy the assumptions of our statistical tests.

```{r log graphs, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

ggplot(Data, aes(x=`Accession`, y=log(`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`), group = factor(Method))) + 
  labs(title="Found SARS-CoV-2 Concentrations for 3 Different Methods - LOG TRANSFORM") +  xlab("Sample") + ylab("log(SARS-CoV-2 Concentration (copies/L))")  + 
  theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
  geom_line(aes(color=Method))+  geom_point(aes(color=Method))+ theme(legend.position='bottom')


```

```{r log all vs all, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

Data4 <- Data3
Data4$'PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)' <- log(Data4$'PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)')
AllvsAll<- Data4 %>% 
  pivot_wider(names_from=c('Method'), 
              values_from='PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)') 

AllvsAllfiltered <- AllvsAll %>% filter(Accession !='23WW361')
AllvsAllfiltered <- AllvsAllfiltered %>% filter(Accession !='23WW291') #low precision
AllvsAllfiltered <- AllvsAllfiltered %>% filter(Accession !='23WW348') #low precision
AllvsAll <- AllvsAllfiltered

ggplot(AllvsAll, aes(x=`Current`, y=`MVP I Kit`)) + 
  labs(title="Current vs MVP I Kit Concentration Measurements - outliers and unsats dropped")+
  ylab("log(MVP I Kit)") + xlab("log(Current)") + theme_bw() + 
  geom_smooth(method = "lm", se=FALSE, color="black") + 
  geom_point() + stat_regline_equation(label.x = 10, label.y = 12.5)+ stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),label.x = 10, label.y = 12.4)

# Plot
ggplot(AllvsAll, aes(x=`Current`, y=`WW Kit`)) + 
  labs(title="Current vs WW Kit Concentration Measurements - outliers and unsats dropped") +  
  xlab("log(Current)") +   ylab("log(WW Kit)")  + theme_bw() + 
  geom_smooth(method = "lm", se=FALSE, color="black") + 
  geom_point() + stat_regline_equation(label.x = 10, label.y = 12.5)+ stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),label.x = 10, label.y = 12.3)

```

Both kits are found to produce results correlated with those of the current method. With an $R^2$ of 0.72, the WW Kit concentrations seem to have a closer "fit" in the model correlating them to those of the current method. To convert between the methods' found concentrations, an equation is given for the line of best fit on the scatterplot, in which "x" is the x axis of the graph (current method). These equations require a log transform on both x and y, making them more complicated than a strictly linear model.

\pagebreak

## ANOVA

```{r log ANOVA, echo=TRUE, warning=FALSE}

# Model
BigModel <- lm(log(Data$`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`) ~ Data$Method)
summary(BigModel)

```

The model fails to find a significant difference in concentrations between methods.

## Pairwise T Tests

```{r log t tests, echo=TRUE, warning=FALSE}

t.test(AllvsAll$`Current`, AllvsAll$`MVP I Kit`, paired = TRUE, alternative = "two.sided")
t.test(AllvsAll$`Current`, AllvsAll$`WW Kit`, paired = TRUE, alternative = "two.sided")

```

The model fails to find a significant difference between the 2 new methods and the current.
