---
title: "Comparison Between Methods"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  classoption: landscape
header-includes:
- \usepackage{titling}
- \usepackage{wrapfig}
- \usepackage{lipsum}
- \usepackage{pdflscape}
- \pretitle{\begin{center} \includegraphics[width=2in,height=2in]{1200px-MassDPH_svg.png}\LARGE\\}
- \posttitle{\end{center}}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

#Load in Libraries
library(readxl)
library(plyr)
library(dplyr)
library(arsenal)
library(ggplot2)
library(writexl)
library(openxlsx)
library(lubridate) 
library(data.table)
library(kableExtra)
library(tidyr)
library(janitor)
library(scales)
library(kableExtra)
library(tinytex)
library(yaml)
library(ggpubr)
library(cowplot)
library(zoo)
library(formattable)
library(treemap)
library(viridis)
library(paletteer)
library(rlist)
library(magrittr)
library(mosaic)
library(tidyverse)

```

```{r, include = F}

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir= normalizePath('..'))
knitr::opts_chunk$set(error = FALSE)

defOut <- knitr::knit_hooks$get("plot")  # save the default plot hook 
knitr::knit_hooks$set(plot = function(x, options) {  # set new plot hook ...
  x <- defOut(x, options)  # first apply the default hook
  if(!is.null(options$wrapfigure)) {  # then, if option wrapfigure is given ...
    # create the new opening string for the wrapfigure environment ...
    wf <- sprintf("\\begin{wrapfigure}{%s}{%g\\textwidth}", options$wrapfigure[[1]], options$wrapfigure[[2]])
    x  <- gsub("\\begin{figure}", wf, x, fixed = T)  # and replace the default one with it.
    x  <- gsub("{figure}", "{wrapfigure}", x, fixed = T)  # also replace the environment ending
  }
  return(x)
})
```
\pagebreak


```{r setup and summary, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

# Pull in Data
Data<-read_excel(here::here("J:/Wastewater Surveillance/Emily/Comparisons/Methods Comparison","KFKitsPoCData_051823.xlsx")) 

Data <- Data %>% drop_na()

# Table
Data2 <- subset(Data, select = c(Method, Accession, `ddPCR date`, `PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`))

knitr::kable(head(Data2, 30), caption = "Data Preview") %>%
  kable_styling("bordered", position = "center") %>%
  kable_styling(latex_options="scale_down")

```

\pagebreak

# Graph of Concentrations

```{r graphs, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

ggplot(Data, aes(x=`Accession`, y=`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)`, group = factor(Method))) + 
  labs(title="Found SARS-CoV-2 Concentrations for 3 Different Methods") +  xlab("Sample") + ylab("SARS-CoV-2 Concentration (copies/L)")  + 
  theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
  geom_line(aes(color=Method))+  geom_point(aes(color=Method))+ theme(legend.position='bottom')


```

\pagebreak

# Comparisons

We do 2 sets of comparisons, comparing both new methods to the current.

```{r graphs all vs all, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

# Reshape
Data3 <- Data2 %>% select(-`ddPCR date`)
AllvsAll<- Data3 %>% 
  pivot_wider(names_from=c('Method'), 
              values_from='PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)') 


ggplot(AllvsAll, aes(x=`Current`, y=`MVP I Kit`)) + 
  labs(title="Current vs MVP I Kit Concentration Measurements")+
  ylab("MVP I Kit")  + theme_bw() + 
  geom_smooth(method = "lm", se=FALSE, color="black") + 
  geom_point() + stat_regline_equation(label.x = 100, label.y = 240000)+ stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")),label.x = 100, label.y = 225000)

# Plot
ggplot(AllvsAll, aes(x=`Current`, y=`WW Kit`)) + 
  labs(title="Current vs WW Kit Concentration Measurements") +  
  xlab("Current Method") + 
  ylab("WW Kit")  + theme_bw() + 
  geom_smooth(method = "lm", se=FALSE, color="black") + 
  geom_point() + stat_regline_equation(label.x = 100, label.y = 300000) +  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), label.x = 100, label.y = 280000)

```

Both kits are found to produce results strongly or very strongly correlated with those of the current method.

\pagebreak

## ANOVA

```{r ANOVA, echo=TRUE, warning=FALSE}

# Model
BigModel <- lm(Data$`PMMoV-Normalized SARS-CoV-2 Concentration (Copies /L)` ~ Data$Method)
summary(BigModel)


```

With a p = 0.3224, the model fails to find a significant difference in concentrations between methods.

## Pairwise T Tests

```{r t tests, echo=TRUE, warning=FALSE}

t.test(AllvsAll$`Current`, AllvsAll$`MVP I Kit`, paired = TRUE, alternative = "two.sided")
t.test(AllvsAll$`Current`, AllvsAll$`WW Kit`, paired = TRUE, alternative = "two.sided")

```

The model fails to find a significant difference between the 2 new methods and the current.


