---
title: "Wastewater Variants Visualization"
output: html_document
date: '2022-12-29'
---


```{r prep, include = FALSE,, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

#Load in Libraries
library(readxl)
library(plyr)
library(dplyr)
library(arsenal)
library(ggplot2)
library(writexl)
library(openxlsx)
library(lubridate) 
library(data.table)
library(kableExtra)
library(tidyr)
library(janitor)
library(scales)
library(kableExtra)
library(tinytex)
library(yaml)
library(ggpubr)
library(cowplot)
library(zoo)
library(formattable)
library(treemap)
library(viridis)
library(paletteer)
library(rlist)
library(magrittr)
library(mosaic)
library(tidyverse)
```

# Data Import
```{r cleaning run 1, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

# Load in new data
all1<-read_excel("J:/Wastewater Surveillance/Emily/Variant Analysis/Data/Cleaned/All1.xlsx") 
all2<-read_excel("J:/Wastewater Surveillance/Emily/Variant Analysis/Data/Cleaned/All2.xlsx") 
all3<-read_excel("J:/Wastewater Surveillance/Emily/Variant Analysis/Data/Cleaned/All3.xlsx") 
all4<-read_excel("J:/Wastewater Surveillance/Emily/Variant Analysis/Data/Cleaned/All4.xlsx") 
all5<-read_excel("J:/Wastewater Surveillance/Emily/Variant Analysis/Data/Cleaned/All5.xlsx") 
all6<-read_excel("J:/Wastewater Surveillance/Emily/Variant Analysis/Data/Cleaned/All6.xlsx") 

```

# Dirty Lineage Plot
```{r dirty Lineage plot, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

ggplot(all3, aes(fill=Lineage, y= Frequency, x=`Accession Number`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Raw")
#ggsave("confusingplot.png")


```

#Cleaning by Filtering to First 4 Characters 

```{r cleaner Lineage plot, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all3$LineageClean <- substr(all3$Lineage, 1, 4)

ggplot(all3, aes(fill=LineageClean, y= Frequency, x=`Accession Number`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Cleaned Lineage Frequency Plot")
ggsave(as.character(paste0('cleanedvarsplot', Sys.Date(), '.png')))

ggplot(all3, aes(fill=LineageClean, y= Frequency, x=Date)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+ scale_x_date(date_labels="%b %y",date_breaks  ="1 day")+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot Has Dates Now")
ggsave(as.character(paste0('dateplot', Sys.Date(), '.png')))

```

#Select Variants 
```{r select variants 2, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all3$SelectVariants2 <- ifelse(substr(all3$Lineage, 1,7) == "XBB.1.5" | substr(all3$Lineage, 1,7) == "BA.2.75", substr(all3$Lineage, 1, 7), 
                             ifelse(substr(all3$Lineage,1,6) == "BQ.1.1", substr(all3$Lineage,1,6),
                                    ifelse(substr(all3$Lineage,1,4) == "BQ.1" | substr(all3$Lineage,1,4) == "BA.5" | substr(all3$Lineage,1,4) == "BN.1", substr(all3$Lineage,1,4),
                                           ifelse(substr(all3$Lineage,1,3) == "XBB" | substr(all3$Lineage,1,3) == "BNF", substr(all3$Lineage,1,3), "All Other Variants"))))

ggplot(all3, aes(fill=SelectVariants2, y= Frequency, x=`Accession Number`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Sample Accession", subtitle = "Highlighting Select SARS-CoV-2 Lineages") +
  scale_fill_manual(values=c('grey', '#F8766D', '#E7861B', '#C59900', '#5BB300', '#00B8EF', '#AC88FF', '#FF689F', '#DC71FA')) + guides(fill=guide_legend(title="Lineage"))
ggsave(as.character(paste0('selectvariants2-', Sys.Date(), '.png')))

ggplot(all3, aes(fill=SelectVariants2, y= Frequency, x=Date)) + 
  geom_bar(stat='identity', position='fill') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+ scale_x_date(date_labels="%b %d %Y",date_breaks  ="1 day")+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Date", subtitle = "Highlighting select SARS-CoV-2 lineages. Duplicate samples are averaged.") + scale_fill_manual(values=c('grey', '#F8766D', '#E7861B', '#C59900', '#5BB300', '#00B8EF', '#AC88FF', '#FF689F', '#DC71FA')) + guides(fill=guide_legend(title="Lineage"))
ggsave(as.character(paste0('selectvariantsdates2-', Sys.Date(), '.png')))

```

# Select Variants - All Data
```{r everything!, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)

all$SelectVariants2 <- ifelse(substr(all$Lineage, 1,7) == "XBB.1.5" | substr(all$Lineage, 1,7) == "BA.2.75", substr(all$Lineage, 1, 7), 
                             ifelse(substr(all$Lineage,1,6) == "BQ.1.1" |  substr(all$Lineage,1,6) == "CH.1.1" , substr(all$Lineage,1,6),
                                    ifelse(substr(all$Lineage,1,4) == "BQ.1" | substr(all$Lineage,1,4) == "BA.5" | 
                                             substr(all$Lineage,1,4) == "BN.1", substr(all$Lineage,1,4),
                                           ifelse(substr(all$Lineage,1,3) == "XBB" | substr(all$Lineage,1,3) == "BNF", substr(all$Lineage,1,3), "All Other Variants"))))

### to keep x labels clean:
everysecond <- function(x){
  x <- sort(unique(x))
  x[seq(2, length(x), 2)] <- ""
  x
}

ggplot(all, aes(fill=SelectVariants2, y= Frequency, x=`Accession Number`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, size=6)) + theme(legend.key.size = unit(.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Sample Accession", subtitle = "Highlighting Select SARS-CoV-2 Lineages") +
  scale_fill_manual(values=c('grey', '#F8766D', '#E08B00', '#BB9D00',  '#AFA100', '#72B000', '#00B8EF', '#AC88FF', '#F763E0', '#DC71FA')) + guides(fill=guide_legend(title="Lineage")) + 
  scale_x_discrete(labels = everysecond(all$`Accession Number`))
ggsave(as.character(paste0('alldatavariants', Sys.Date(), '.png')))

ggplot(all, aes(fill=SelectVariants2, y= Frequency, x=Date)) + 
  geom_bar(stat='identity', position='fill') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+ scale_x_date(date_labels="%b %d %Y",date_breaks  ="1 week")+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Date", subtitle = "Highlighting select SARS-CoV-2 lineages. Duplicate samples are averaged.") + scale_fill_manual(values=c('grey', '#F8766D', '#E08B00', '#BB9D00',  '#AFA100', '#72B000', '#00B8EF', '#AC88FF', '#F763E0', '#DC71FA')) + guides(fill=guide_legend(title="Lineage"))
ggsave(as.character(paste0('alldatavariantsdates', Sys.Date(), '.png')))

```

```{r everything averaged by week, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all$week <- isoweek(all$Date)
all$year <-year(all$Date)

allbyweek <- all%>% aggregate(Frequency ~ week + year + SelectVariants2, FUN = mean)

allbyweek$date <- as.Date(lubridate::parse_date_time(paste(allbyweek$year, allbyweek$week, 1, sep="/"),'Y/W/w'))

ggplot(allbyweek, aes(fill=SelectVariants2, y= Frequency, x=date)) + 
  geom_bar(stat='identity', position='fill') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+ scale_x_date(date_labels="%b %d %Y",date_breaks  ="1 week")+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Date", subtitle = "Highlighting select SARS-CoV-2 lineages. Duplicate samples are averaged.") + scale_fill_manual(values=c('grey', '#F8766D', '#E08B00', '#BB9D00',  '#AFA100', '#72B000', '#00B8EF', '#AC88FF', '#F763E0', '#DC71FA')) + guides(fill=guide_legend(title="Lineage"))
ggsave(as.character(paste0('alldatavariantsweeks', Sys.Date(), '.png')))


```



# Select Variants - Meagan
```{r everything!, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)
view(all)

### Collapse silly genetics rules: BF-not-BF.7 and BE are BA.5s
all$Lineage2 <- ifelse(substr(all$Lineage, 1,2) == "BF" & 
                                     substr(all$Lineage, 1, 4) != "BF.7", "BA.5",                          #BF-not-BF.7s
                                     ifelse(substr(all$Lineage, 1,2) == "BE", "BA.5",                      #BEs
                         all$Lineage))

### Collapse to substrings by grepping, in order to be net and tidy
all$SelectVariants2 <- ifelse(substr(all$Lineage, 1,7) == "XBB.1.5", substr(all$Lineage, 1, 7), 
                             ifelse(substr(all$Lineage,1,6) == "BQ.1.1" |  substr(all$Lineage,1,6) == "CH.1.1" , substr(all$Lineage,1,6),
                                    ifelse(substr(all$Lineage,1,4) == "BQ.1" | substr(all$Lineage,1,4) == "BA.5" | 
                                             substr(all$Lineage,1,4) == "BN.1" | substr(all$Lineage,1,4) == "BF.7", substr(all$Lineage,1,4),
                                           ifelse(substr(all$Lineage,1,3) == "XBB", substr(all$Lineage,1,3), "All Other Variants"))))

ggplot(all, aes(fill=SelectVariants2, y= Frequency, x=`Accession Number`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, size=6)) + theme(legend.key.size = unit(.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Sample Accession", subtitle = "Highlighting Select SARS-CoV-2 Lineages") +scale_fill_manual(values=c('grey', '#F8766D', '#E08B00', '#BB9D00', '#72B000', '#00B8EF', "7997FF", '#AC88FF', '#F763E0', '#DC71FA')) + guides(fill=guide_legend(title="Lineage")) + 
  scale_x_discrete(labels = everysecond(all$`Accession Number`))
ggsave(as.character(paste0('alldatavariantsmeagan', Sys.Date(), '.png')))

ggplot(all, aes(fill=SelectVariants2, y= Frequency, x=Date)) + 
  geom_bar(stat='identity', position='fill') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+ scale_x_date(date_labels="%b %d %Y",date_breaks  ="1 week")+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Date", subtitle = "Highlighting select SARS-CoV-2 lineages. Duplicate samples are averaged.") + scale_fill_manual(values=c('grey', '#F8766D', '#E08B00', '#BB9D00', '#72B000', '#00B8EF', "7997FF", '#AC88FF', '#F763E0', '#DC71FA')) + guides(fill=guide_legend(title="Lineage"))
ggsave(as.character(paste0('alldatavariantsdatesmeagan', Sys.Date(), '.png')))

all$week <- isoweek(all$Date)
all$year <-year(all$Date)

allbyweek <- all%>% aggregate(Frequency ~ week + year + SelectVariants2, FUN = mean)

allbyweek$date <- as.Date(lubridate::parse_date_time(paste(allbyweek$year, allbyweek$week, 1, sep="/"),'Y/W/w'))

ggplot(allbyweek, aes(fill=SelectVariants2, y= Frequency, x=date)) + 
  geom_bar(stat='identity', position='fill') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(.4, 'cm'))+ scale_x_date(date_labels="%b %d %Y",date_breaks  ="1 week")+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot by Week", subtitle = "Highlighting select SARS-CoV-2 lineages. Duplicate samples are averaged.") + scale_fill_manual(values=c('grey', '#F8766D', '#E08B00', '#BB9D00', '#72B000', '#00B8EF', "7997FF", '#AC88FF', '#F763E0', '#DC71FA')) + guides(fill=guide_legend(title="Lineage"))
ggsave(as.character(paste0('alldatavariantsweeks', Sys.Date(), '.png')))

```


#Luc's Experimental Scheme

```{r luc family tree condensing format, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

### Luc suggested a way to collapse things down: if a lineage has a frequency of under 1% and contains a decimal, lop off the last decimal and stuff after it to categorize with the parent lineage. For example, if BA.1.1.1 is 0.09%, then cut it to BA.1.1. Stop once all lineages have over 1% frequency in order to aid in visualization
all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)

all$Infrequent <- ifelse(all$Frequency < 0.01, 0, 1)
all$SelectVariants3 <- ifelse(all$Infrequent == "0" & is.na(as.numeric(str_sub(all$Lineage, start= -3)))!=TRUE, str_sub(all$Lineage, end= -3), 
                              ifelse(all$Infrequent == "0" & is.na(as.numeric(str_sub(all$Lineage, start= -2)))!=TRUE, str_sub(all$Lineage, end= -2), all$Lineage))

all$SelectVariants3 <- ifelse(str_sub(all$SelectVariants3, start= -1)==".", substr(all$SelectVariants3, 1, nchar(all$SelectVariants3)-1), all$SelectVariants3)
view(all)
forluc <- all %>% dplyr::select(`Accession Number`, Date, SelectVariants3, Frequency)
forluc_sum1 <- aggregate(Frequency ~ `Accession Number` + Date + SelectVariants3, forluc, sum)

#Repeating this
forluc_sum <- forluc_sum1
forluc_sum$Infrequent <- ifelse(forluc_sum$Frequency < 0.01, 0, 1)
forluc_sum$SelectVariants3 <- ifelse(forluc_sum$Infrequent == "0" & is.na(as.numeric(str_sub(forluc_sum$SelectVariants3, start= -3)))!=TRUE, str_sub(forluc_sum$SelectVariants3, end= -3), 
                              ifelse(forluc_sum$Infrequent == "0" & is.na(as.numeric(str_sub(forluc_sum$SelectVariants3, start= -2)))!=TRUE, str_sub(forluc_sum$SelectVariants3, end= -2), forluc_sum$SelectVariants3))

forluc_sum$SelectVariants3 <- ifelse(str_sub(forluc_sum$SelectVariants3, start= -1)==".", substr(forluc_sum$SelectVariants3, 1, nchar(forluc_sum$SelectVariants3)-1), forluc_sum$SelectVariants3)
forluc <- all %>% dplyr::select(`Accession Number`, Date, SelectVariants3, Frequency)
forluc_sum <- aggregate(Frequency ~ `Accession Number` + Date + SelectVariants3, forluc, sum)



view(forluc_sum)
ggplot(forluc_sum, aes(fill=SelectVariants3, y= Frequency, x=`Accession Number`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Luc's Method")

ggplot(forluc_sum, aes(fill=SelectVariants3, y= Frequency, x=Date)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.4, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Luc's Method")


```
# DIY Variant Selection

## Regression Model
```{r DIY variant selection, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)
all$Date <- as.Date(all$Date, format="%Y-%m-%d")
all <- all %>% select(Lineage, Date, Frequency)
all <- all %>% 
  group_by(Lineage, Date) %>% 
  summarize(avg=mean(Frequency)) # get rid of duplicate samples of same day - average them
#all <- allb %>% filter(Date > as.Date("2022-12-25", format="%Y-%m-%d"))
all$FrequencyPerc <- all$avg*100
all$LineageClean <- as.factor(all$Lineage)
all$Date <- as.Date(all$Date, format="%Y-%m-%d")

hist(all$avg)
#ofinterestmodel <- all %>% 
#   arrange(desc(mdy(Date))) %>%
#   group_by(`LineageClean`) %>% 
#   slice(tail(row_number(), 7))%>% ungroup #chop down all date to get last 7 (or fewer) data points of each raw variant

#model the frequency vs date by lineage
library(betareg)
model1 <- betareg(avg ~ Date + LineageClean, data=all)
summary(model1)
model1coef <- summary(model1)$coefficients #get out coefficients
model1coef <- tibble::rownames_to_column(as.data.frame(model1coef), "row_names") # Apply rownames_to_column
summary(model1coef)

#generate list of significantly changing variants (both pos and neg for now)
changing <- model1coef %>% filter(`mean.Pr...z..`< 0.05)
changing$Lineage <- str_remove(changing$row_names, "LineageClean")



library(broom)
m1_res <- augment(model1, data = all)
ggplot(m1_res, aes(x = Date, y = .resid)) +
        geom_point() +
        stat_smooth(method = "loess", formula = "y ~ x")  +
        labs(y = "Residuals", x = "Date",
             title = "Residuals as a Function of Date")


#merge to mark the "all" dataset
changing$growthflag <- "CHANGING"
all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)
allmodel<- merge(x = all, y = changing, all.x = TRUE) #merge it with the main data - now the data are flagged "growing" or no
allmodel$growthvariants <- ifelse(allmodel$growthflag=="CHANGING", allmodel$Lineage, #dump non-growing vars into an NA category
                             NA)
ggplot(allmodel, aes(fill=growthvariants, y= Frequency, x=`Date`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Using Beta Regression Model")
```

## Arithmetic - Last 7 Appearances Per Variant

### 4+ Increasing Days
```{r DIY variant selection arithmetic, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)

#Some data rearranging
all$Date <- as.Date(all$Date, format="%Y-%m-%d")
all <- all %>% select(Lineage, Date, Frequency)
allb <- all %>% 
  group_by(Lineage, Date) %>% 
  summarize(avg=mean(Frequency)) # get rid of duplicate samples of same day - average them
latest <- allb %>% 
   arrange(desc(mdy(Date))) %>%
   group_by(`Lineage`) %>% 
   slice(tail(row_number(), 7))%>% ungroup #chop down all date to get last 7 (or fewer) data points of each raw variant
latest <- latest %>% 
   arrange(Date) %>%
   group_by(Lineage) %>% 
   mutate(Index=1:n()) #number each variant+day combo (index = Nth appearance of [name of variant])=
latest <- latest %>% select(Lineage, Index, avg) #drop the dates - we don't need them
dailies <- spread(latest, key = Index, value = avg) #widen

#Find differences in frequency between 1st and 2nd appearance, 2nd and 3rd, etc
dailies$change1to2 <- dailies$`2` - dailies$`1` 
dailies$change2to3 <- dailies$`3` - dailies$`2` 
dailies$change3to4 <- dailies$`4` - dailies$`2` 
dailies$change4to5 <- dailies$`5` - dailies$`4` 
dailies$change5to6 <- dailies$`6` - dailies$`5` 
dailies$change6to7 <- dailies$`7` - dailies$`6` 

#Now we flag these: number of NAs, number of increases, number of decreases
dailies$na_count <- rowSums(is.na(dailies[c('change1to2', 'change2to3', 'change3to4', 'change4to5', 'change5to6','change6to7')])) 
dailies$poscount <- rowSums(dailies[c('change1to2', 'change2to3', 'change3to4', 'change4to5', 'change5to6','change6to7')] >0)
dailies$negcount <- rowSums(dailies[c('change1to2', 'change2to3', 'change3to4', 'change4to5', 'change5to6','change6to7')] <0)

#That was shockingly easy. Now we flag variants as increasing
dailies$growthflag <- ifelse(dailies$poscount >= 4, "GROWING", NA)

#Clean this up - pull out just the name of the variant and the flag
growthflag <- dailies %>% select(Lineage, growthflag)

#Merge with the all-data dataset once more:
all<- merge(x = all, y = growthflag, all.x = TRUE)
all$growthvariants <- ifelse(all$growthflag=="GROWING", all$Lineage, 
                             "Other")

ggplot(all, aes(fill=growthvariants, y= Frequency, x=`Date`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Raw")
```

### SD Threshold
```{r DIY variant selection SD, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

#Uses whichever "dailies" is used last - if you want the first 7 appearances of every variant, go back and run that chunk 

forhist <- gather(dailies, cat, measurement, 9:14, factor_key=TRUE) # I want to check that the changes are distributed normally
hist(forhist$measurement, breaks = 50) #oh they are! nice

dailies$avgchange <- rowMeans(dailies[,9:14], na.rm=TRUE) #across each row (ie for each variant) get mean of all freq changes 
SD <- sd(dailies$avgchange, na.rm = TRUE) #SD of these average changes
SD2 <- sd(unlist(dailies[,9:14]), na.rm=TRUE) #SD of stepwise changes
meanchange <- mean(dailies$avgchange, na.rm=TRUE) #mean of all changes

SD2
meanchange


dailies$growthflag2 <- ifelse(dailies$avgchange >= (meanchange + 0.4*SD2), "GROWING", NA) #call variant growing if avg change of variant > mean + n*SD
growthflag2 <- dailies %>% select(Lineage, growthflag2) #pull out this info
allsd<- merge(x = all, y = growthflag2, all.x = TRUE) #merge it with the main data - now the data are flagged "growing" or no

allsd$growthvariants <- ifelse(allsd$growthflag2=="GROWING", allsd$Lineage, #dump non-growing vars into an NA category
                             NA)

ggplot(allsd, aes(fill=growthvariants, y= Frequency, x=`Date`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Raw")
```

## Arithmetic - Last 7 Days of Data

### 4+ Increasing Days
```{r DIY variant selection arithmetic 2, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}
 
# Repeating the above, just not with last 7 appearances - filter to the past 7 DAYS of data.
all <- rbind(all2, all3raw, all4raw, all5raw, all6raw)

#Some data rearranging
all$Date <- as.Date(all$Date, format="%Y-%m-%d")
all <- all %>% select(Lineage, Date, Frequency)
allb <- all %>% 
  group_by(Lineage, Date) %>% 
  summarize(avg=mean(Frequency)) # get rid of duplicate samples of same day - average them
latest <- allb %>% filter(Date > as.Date("2022-12-25", format="%Y-%m-%d"))
latest <- latest %>% 
   arrange(Date) %>%
   group_by(Lineage) %>% 
   mutate(Index=1:n()) #number each variant+day combo (index = Nth appearance of [name of variant])=
latest <- latest %>% select(Lineage, Index, avg) #drop the dates - we don't need them
dailies <- spread(latest, key = Index, value = avg) #widen

#Find differences in frequency between 1st and 2nd appearance, 2nd and 3rd, etc
dailies$change1to2 <- dailies$`2` - dailies$`1` 
dailies$change2to3 <- dailies$`3` - dailies$`2` 
dailies$change3to4 <- dailies$`4` - dailies$`2` 
dailies$change4to5 <- dailies$`5` - dailies$`4` 
dailies$change5to6 <- dailies$`6` - dailies$`5` 
dailies$change6to7 <- dailies$`7` - dailies$`6` 

#Now we flag these: number of NAs, number of increases, number of decreases
dailies$na_count <- rowSums(is.na(dailies[c('change1to2', 'change2to3', 'change3to4', 'change4to5', 'change5to6','change6to7')])) 
dailies$poscount <- rowSums(dailies[c('change1to2', 'change2to3', 'change3to4', 'change4to5', 'change5to6','change6to7')] >0)
dailies$negcount <- rowSums(dailies[c('change1to2', 'change2to3', 'change3to4', 'change4to5', 'change5to6','change6to7')] <0)

#That was shockingly easy. Now we flag variants as increasing
dailies$growthflag <- ifelse(dailies$poscount >= 4, "GROWING", NA)

#Clean this up - pull out just the name of the variant and the flag
growthflag <- dailies %>% select(Lineage, growthflag)

#Merge with the all-data dataset once more:
all<- merge(x = all, y = growthflag, all.x = TRUE)

all$growthvariants <- ifelse(all$growthflag=="GROWING", all$Lineage, 
                             "Other")

ggplot(all, aes(fill=growthvariants, y= Frequency, x=`Date`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Raw")
```
### SD Threshold
```{r DIY variant selection SD 2, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}

forhist <- gather(dailies, cat, measurement, 9:14, factor_key=TRUE) # I want to check that the changes are distributed normally
hist(forhist$measurement, breaks = 50) #oh they are! nice

dailies$avgchange <- rowMeans(dailies[,9:14], na.rm=TRUE) #across each row (ie for each variant) get mean of all freq changes 
SD <- sd(dailies$avgchange, na.rm = TRUE) #SD of these average changes
SD2 <- sd(unlist(dailies[,9:14]), na.rm=TRUE) #SD of stepwise changes
meanchange <- mean(dailies$avgchange, na.rm=TRUE) #mean of all changes

SD2
meanchange


dailies$growthflag2 <- ifelse(dailies$avgchange >= (meanchange + 0.4*SD2), "GROWING", NA) #call variant growing if avg change of variant > mean + n*SD
growthflag2 <- dailies %>% select(Lineage, growthflag2) #pull out this info
allsd<- merge(x = all, y = growthflag2, all.x = TRUE) #merge it with the main data - now the data are flagged "growing" or no

allsd$growthvariants <- ifelse(allsd$growthflag2=="GROWING", allsd$Lineage, #dump non-growing vars into an NA category
                             NA)

ggplot(allsd, aes(fill=growthvariants, y= Frequency, x=`Date`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Raw")
```

#Working with Estimated Variant Concentrations

```{r estimated conc, echo=FALSE , warning=FALSE, message=FALSE, results = "asis", type = 'latex'}
 
#Get all the more recent data together
all <- rbind(all2, all3, all4raw, all5raw, all6raw)

#Turn percents into concentrations
all$`Estimated Concentration` <- all$`Concentration (copies/L)`*as.numeric(all$Frequency)

#Now what if I plot these?
ggplot(all, aes(x=`Date`, y=`Estimated Concentration`, color=`Lineage`)) +
  geom_point() + theme(legend.position="none")

#Examine Distrib
plot(all$Date, log(all$`Concentration (copies/L)`))
histogram(all$`Estimated Concentration`) #right skew ew
histogram(sqrt(all$`Estimated Concentration`))
histogram(log(all$`Estimated Concentration`)) #perfect!

#Model the concentration vs date by lineage
model1 <- lm(log(`Estimated Concentration`) ~ Date + Lineage, data=all)
summary(model1)
model1coef <- summary(model1)$coefficients #get out coefficients
model1coef <- tibble::rownames_to_column(as.data.frame(model1coef), "row_names") # Apply rownames_to_column
summary(model1coef)

#generate list of significantly changing variants (both pos and neg for now)
changing <- model1coef %>% filter(`Pr(>|t|)`< 0.1)
changing$Lineage <- str_remove(changing$row_names, "Lineage")

m1_res <- augment(model1, data = all)
ggplot(m1_res, aes(x = Date, y = .resid)) +
        geom_point() +
        stat_smooth(method = "loess", formula = "y ~ x")  +
        labs(y = "Residuals", x = "Date",
             title = "Residuals as a Function of Date")


#merge to mark the "all" dataset
changing$growthflag <- "CHANGING"
allmodel<- merge(x = all, y = changing, all.x = TRUE) #merge it with the main data - now the data are flagged "growing" or no
allmodel$growthvariants <- ifelse(allmodel$growthflag=="CHANGING", allmodel$Lineage, #dump non-growing vars into an NA category
                             NA)
XBB <- all %>% filter(Lineage == "XBB.1.5")
plot(XBB$Date, log(XBB$`Estimated Concentration`))

BQ <- all %>% filter(Lineage == "BQ.1.1")
plot(BQ$Date, log(BQ$`Estimated Concentration`))

BA2 <- all %>% filter(Lineage == "BA.2.75")
plot(BA2$Date, log(BA2$`Estimated Concentration`))

ggplot(allmodel, aes(fill=growthvariants, y= Frequency, x=`Date`)) + 
  geom_bar(position='fill', stat='identity') + theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) + theme(legend.key.size = unit(0.1, 'cm'))+
  scale_y_continuous(limits = c(0, 1)) + labs(color = "Variants", title = "Lineage Frequency Plot - Using Linear Regression Model on Concentraiton Data")
```
